---
title: "Distance to Ocean"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

-   In ArcGIS Pro v3.1.1., an 'empty' grid was generated that covers the extent of the SCM region of interest. The grid was created using the `Create Fishnet` tool with the following parameters:
    -   Extent: SCM region boundary
    -   Cell size width: 100 ft
    -   Cell size height: 100 ft
-   The Euclidean distance from the centroid of each grid cell to the [Pacific Ocean](https://geodata.library.ucsb.edu/catalog/3853-s3_2002_s3_reg_pacific_ocean) was calculated using the `Near` tool
-   This Quarto doc rasterizes the Euclidean distance grid, preserving the value represeting the distance from each grid cell to the ocean, to a 32-bit floating point raster.

## Clear environment

```{r}
rm(list = ls())
```

## Load packages

```{r}
library(terra)
library(tidyverse)
library(here)
library(sf)
```

## Load data

```{r}
# canopy height model (45ft)
scm_chm_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_chm_45ft.tif")
crs(scm_chm_45ft)

# distance to nearest river points 
dist_ocean_points <- read_sf("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/distance_to_oceans/distance_to_oceans_points.shp")
crs(dist_ocean_points)

# scm region boudary
scm_region_boundary <- read_sf("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/boundaries/scm_region/scm_region_boundary.shp")
crs(scm_region_boundary)
```

## Turn the ocean points into a raster

#### Rasterize the distance to ocean points created in ArcGIS Pro

```{r}
# reproject dist_oceans_points to the scm_dem_45ft projection
dist_oceans_points_proj <- st_transform(dist_ocean_points, 
                                        crs(scm_chm_45ft))

# make sure both are in same projection 
crs(scm_chm_45ft) == crs(dist_oceans_points_proj)
st_crs(dist_oceans_points_proj)$units #check units


# Create a new (blank) raster
new_raster <- raster()
# match raster extent to that of the hec data
extent(new_raster) <- extent(dist_oceans_points_proj)
crs(new_raster) <- crs(dist_oceans_points_proj) # assign new raster same CRS
res(new_raster) <- 100 # match resolution to pixel size 

# rasterize and preserve the distance to ocean values
dist_oceans_raster <- rasterize(dist_oceans_points_proj,
                                new_raster, 
                                field = "ocean_dist")

# plot the raster
plot(dist_oceans_raster)

# # save the raster
# writeRaster(dist_oceans_raster, 
#             "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/distance_to_oceans/distance_to_oceans_res100ft.tif",
#             overwrite = TRUE)
```

#### Resample the ocean raster to the DEM resolution

```{r}
# load the distance to rivers raster as a SpatRaster
# the distance to rivers is in meters
dist_oceans_raster <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/distance_to_oceans/distance_to_oceans_res100ft.tif")

# confirm they're in the same CRS
crs(scm_chm_45ft) == crs(dist_oceans_raster)

# resample the river raster to the DEM resolution
dist_oceans_raster_45ft <- resample(dist_oceans_raster, 
                                    scm_chm_45ft, 
                                    method = "bilinear")

# check the resolution
res(dist_oceans_raster_45ft) == res(scm_chm_45ft)
plot(dist_oceans_raster_45ft)

# Apply focal function to calculate the mean of neighboring cells, only for NAs
dist_oceans_raster_45ft_filled <- ifel(is.na(dist_oceans_raster_45ft), 
                                       focal(dist_oceans_raster_45ft, 
                                             w = 3, fun = mean, 
                                             na.rm = TRUE), 
                                       dist_oceans_raster_45ft)


# plot the resampled river raster
plot(dist_oceans_raster_45ft_filled)


# apply the focal function again to the filled raster
# STOP: this may not be necessary; the function is just a little buggy and 
# didn't fill every cell the first time
dist_oceans_raster_45ft_filled <- ifel(is.na(dist_oceans_raster_45ft_filled), 
                                       focal(dist_oceans_raster_45ft_filled, 
                                             w = 3, fun = mean, 
                                             na.rm = TRUE), 
                                       dist_oceans_raster_45ft_filled)

# plot the filled river raster
plot(dist_oceans_raster_45ft_filled)

# mask and crop the rivers raster
dist_oceans_raster_45ft_filled <- mask(dist_oceans_raster_45ft_filled, 
                                       scm_region_boundary) %>% 
  crop(scm_region_boundary)
plot(dist_oceans_raster_45ft_filled)

# save the resampled river raster
dist_ocean_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/distance_to_oceans/distance_to_oceans_res45ft.tif"
writeRaster(dist_oceans_raster_45ft_filled, dist_ocean_path, 
            overwrite = TRUE)
```
