---
title: "ArcGIS Pro Land Prioritization"
author: "Maxwell Pepperdine"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

This document outlines the process of using ArcGIS Pro to prioritize land parcels for conservation, and then visualized the results. All county parcels (protected & unprotected) in the SCM region AND only unprotected parcels in the SCM region with the greatest area of suitable habitat were identified by the following steps:

- Make sure the data is in the correct projection (NAD 83 CA Zone III)
- Intersect the post-fire suitable habitat with the SCM region parcels 
- Dissolve by APN to get total area of suitable habitat per parcel
- Use the calcuate geometry tool to get the area of suitable habitat (in hectares and square meters) within each parcel

## Clear environment

```{r}
rm(list=ls())
```

## Load packages

```{r}
library(terra)
library(sf)
library(tidyverse)
library(patchwork)
```


## Load Data

```{r}
# all SCM parcels with the area of suitable habitat
all_scm_parcels_suit_hab <- st_read("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/data/habitat_suitability/parcel_analysis/all_parcel_postfire_suit_hab_area.shp")

# unprotected SCM parcels with the area of suitable habitat
unprotected_scm_parcels_suit_hab <- st_read("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/data/habitat_suitability/parcel_analysis/unprotected_parcel_postfire_suit_hab_area.shp")
```

## Clean the data

```{r}
# all SCM parcels
all_scm_parcels_suit_hab <- all_scm_parcels_suit_hab %>% 
  select(APN, 
         suit_area_m2 = suitable_a, 
         suit_area_ha = suit_area_)

# unprotected SCM parcels
unprotected_scm_parcels_suit_hab <- unprotected_scm_parcels_suit_hab %>% 
  select(APN, 
         suit_area_m2 = suit_hab_m, 
         suit_area_ha = suit_hab_h)
```

## Select the top 10 parcels with the greatest area of suitable habitat

```{r}
# all SCM parcels
top_10_all_scm_parcels <- all_scm_parcels_suit_hab %>% 
  arrange(desc(suit_area_ha)) %>% 
  head(10)
# # save the top 10 parcels
# st_write(top_10_all_scm_parcels, 
#          "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/data/habitat_suitability/parcel_analysis/top_10_all_scm_parcels.shp")


# unprotected SCM parcels
top_10_unprotected_scm_parcels <- unprotected_scm_parcels_suit_hab %>% 
  arrange(desc(suit_area_ha)) %>% 
  head(10)
# save the top 10 parcels
# st_write(top_10_unprotected_scm_parcels, 
#          "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/data/habitat_suitability/parcel_analysis/top_10_unprotected_scm_parcels.shp")
```

## Make some plots

```{r}
# all SCM parcels
all_scm_plot <- ggplot(top_10_all_scm_parcels, 
                       aes(x = reorder(APN, suit_area_ha), 
                           y = suit_area_ha, 
                           fill = suit_area_ha)) +
  geom_col(color = "black", 
           linewidth = 0.2) +
  scale_fill_gradient(low = "grey80", 
                      high = "lightgreen") +
  labs(title = " ",
       x = "APN",
       y = "Area of suitable habitat (ha)", 
       fill = "APN") + 
  theme_classic() + 
  geom_text(aes(label = round(suit_area_ha, 0)), 
            vjust = -0.5, 
            size = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
#all_scm_plot

# unprotected SCM parcels
unprotected_scm_plot <- ggplot(top_10_unprotected_scm_parcels, 
                               aes(x = reorder(APN, suit_area_ha), 
                                   y = suit_area_ha, 
                                   fill = suit_area_ha)) +
  geom_col(color = "black", 
           linewidth = 0.2) +
  scale_fill_gradient(low = "grey80", 
                      high = "lightgreen") +
  labs(title = " ",
       x = "APN",
       y = " ", 
       fill = " ") + 
  theme_classic() + 
  geom_text(aes(label = round(suit_area_ha, 0)), 
            vjust = -0.5, 
            size = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
#unprotected_scm_plot

combined_plot <- all_scm_plot / unprotected_scm_plot 
combined_plot <- combined_plot + plot_annotation(tag_levels = "A")
#combined_plot

combined_plot1 <- all_scm_plot + unprotected_scm_plot
combined_plot1 <- combined_plot1 + plot_annotation(tag_levels = "A")
combined_plot1

# save the plots
# ggsave("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/layouts/parcel_prioritization.png", 
#        plot = combined_plot, 
#        width = 10, 
#        height = 10, 
#        units = "in")

# ggsave("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/layouts/parcel_prioritization1.png", 
#        plot = combined_plot1, 
#        width = 10, 
#        height = 6, 
#        units = "in")
```

```{r}
# all SCM parcels
all_scm_plot <- ggplot(top_10_all_scm_parcels, 
                       aes(x = reorder(APN, suit_area_ha), 
                           y = suit_area_ha, 
                           fill = suit_area_ha)) +
  geom_col(color = "black", 
           linewidth = 0.2) +
  scale_fill_gradient(low = "grey80", 
                      high = "lightgreen") +
  labs(title = " ",
       x = "APN",
       y = "Area of suitable habitat (ha)", 
       fill = "APN") + 
  theme_classic() + 
  geom_text(aes(label = round(suit_area_ha, 0)), 
            vjust = -0.5, 
            size = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
#all_scm_plot

# unprotected SCM parcels
unprotected_scm_plot <- ggplot(top_10_unprotected_scm_parcels, 
                               aes(x = reorder(APN, suit_area_ha), 
                                   y = suit_area_ha, 
                                   fill = suit_area_ha)) +
  geom_col(color = "black", 
           linewidth = 0.2) +
  scale_fill_gradient(low = "grey80", 
                      high = "dodgerblue3") +
  labs(title = " ",
       x = "APN",
       y = " ", 
       fill = " ") + 
  theme_classic() + 
  geom_text(aes(label = round(suit_area_ha, 0)), 
            vjust = -0.5, 
            size = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none")
#unprotected_scm_plot


combined_plot2 <- all_scm_plot + unprotected_scm_plot
combined_plot2 <- combined_plot2 + plot_annotation(tag_levels = "A")
combined_plot2


ggsave("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/outputs/layouts/parcel_prioritization2.png",
       plot = combined_plot2,
       width = 10,
       height = 6,
       units = "in")
```















 
