---
title: "Prepping Env Variables for Wallace"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

-   `Wallace` requires that all environmental predictor rasters are in the same projection, extent, and resolution; this document will ensure that.
-   Occurrence data uploaded to `Wallace` will be in WGS 84 by default, so all environmental predictor rasters need to be projected to WGS 84.

## Clear history

```{r}
rm(list = ls())
```

## Load packages

```{r}
library(wallace)
library(terra)
library(sf)
library(tidyverse)
```

## Make sure all rasters being used in `Wallace` are in the same projection, extent, resolution

#### Load environmental predictor data

```{r}
# CHM 
chm_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_chm_45ft.tif"
scm_chm_45ft <- rast(chm_file_path)

# Distance to rivers
dist_to_rivers_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/distance_to_rivers_res45ft.tif"
dist_to_rivers <- rast(dist_to_rivers_file_path)

# Distance to oceans
dist_ocean_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/distance_to_oceans_res45ft.tif"
dist_to_oceans <- rast(dist_ocean_path)

# DEM 
dem_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_dem_45ft.tif"
scm_dem_45ft <- rast(dem_file_path)

# Aspect 
aspect_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_aspect_45ft.tif"
scm_aspect_45ft <- rast(aspect_file_path)
```

#### Check the projection, resolution, and extent of the rasters

```{r}
# make sure the projection, resolution, and extent of all rasters are the same

## projection 
crs(scm_chm_45ft) == crs(scm_dem_45ft)
crs(scm_chm_45ft) == crs(scm_aspect_45ft)
crs(scm_chm_45ft) == crs(dist_to_rivers)
crs(scm_chm_45ft) == crs(dist_to_oceans)

## extent
ext(scm_chm_45ft) == ext(scm_dem_45ft)
ext(scm_chm_45ft) == ext(scm_aspect_45ft)
ext(scm_chm_45ft) == ext(dist_to_rivers)
ext(scm_chm_45ft) == ext(dist_to_oceans)

## resolution
res(scm_chm_45ft) == res(scm_dem_45ft)
res(scm_chm_45ft) == res(scm_aspect_45ft)
res(scm_chm_45ft) == res(dist_to_rivers)
res(scm_chm_45ft) == res(dist_to_oceans)
```

#### Resample the DEM and Aspect rasters to the extent of the CHM

These two rasters didn't have the exact same extent at the CHM. This will cause issues when running Maxent in `Wallace`, so we need to resample them to the same extent.

```{r}
# the dem and aspect rasters are not the same extent as the chm
# resample them to the extent of the chm
scm_dem_45ft_resample <- resample(scm_dem_45ft, 
                                  scm_chm_45ft)
scm_aspect_45ft_resample <- resample(scm_aspect_45ft, 
                                     scm_chm_45ft)

# check the resampled rasters to make sure everything worked
crs(scm_chm_45ft) == crs(scm_dem_45ft_resample)
crs(scm_chm_45ft) == crs(scm_aspect_45ft_resample)
crs(scm_chm_45ft) == crs(dist_to_rivers)

ext(scm_chm_45ft) == ext(scm_dem_45ft_resample)
ext(scm_chm_45ft) == ext(scm_aspect_45ft_resample)
ext(scm_chm_45ft) == ext(dist_to_rivers)

res(scm_chm_45ft) == res(scm_dem_45ft_resample)
res(scm_chm_45ft) == res(scm_aspect_45ft_resample)
res(scm_chm_45ft) == res(dist_to_rivers)

# save the resampled rasters
writeRaster(scm_dem_45ft_resample, 
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_dem_45ft_resample.tif", 
            overwrite = TRUE)
writeRaster(scm_aspect_45ft_resample,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_aspect_45ft_resample.tif",
            overwrite = TRUE)
```

#### Put all the rasters in WGS 84

```{r}
# project the scm_chm_45ft raster to WGS 84
scm_chm_wgs84 <- project(scm_chm_45ft, 
                         "EPSG:4326")
crs(scm_chm_wgs84) # QC

# load all the other rasters
scm_chm_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_chm_45ft.tif")
scm_dem_45ft_resample <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_dem_45ft_resample.tif")
scm_aspect_45ft_resample <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_aspect_45ft_resample.tif")
dist_to_rivers <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/distance_to_rivers_res45ft.tif")
dist_to_oceans <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/distance_to_oceans_res45ft.tif")
scm_postfire_chm_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_postfire_chm_45ft.tif")


# change the CRS of all these rasters to WGS 84
scm_chm_45ft_wgs84 <- project(scm_chm_45ft, 
                              crs(scm_chm_wgs84))
scm_dem_45ft_resample_wgs84 <- project(scm_dem_45ft_resample, 
                                      crs(scm_chm_wgs84))
scm_aspect_45ft_resample_wgs84 <- project(scm_aspect_45ft_resample, 
                                         crs(scm_chm_wgs84))
dist_to_rivers_wgs84 <- project(dist_to_rivers, 
                               crs(scm_chm_wgs84))
dist_to_oceans_wgs84 <- project(dist_to_oceans, 
                               crs(scm_chm_wgs84))
scm_postfire_chm_45ft_wgs84 <- project(scm_postfire_chm_45ft, 
                                      crs(scm_chm_wgs84))
```

#### Re-check the projection, resolution, and extent of the rasters

```{r}
scm_chm_45ft_wgs84 <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/scm_chm_wgs84.tif")

# projection
crs(scm_chm_45ft_wgs84) == crs(scm_dem_45ft_resample_wgs84)
crs(scm_chm_45ft_wgs84) == crs(scm_aspect_45ft_resample_wgs84)
crs(scm_chm_45ft_wgs84) == crs(dist_to_rivers_wgs84)
crs(scm_chm_45ft_wgs84) == crs(dist_to_oceans_wgs84)
crs(scm_chm_45ft_wgs84) == crs(scm_postfire_chm_45ft_wgs84)

# extent
ext(scm_chm_45ft_wgs84) == ext(scm_dem_45ft_resample_wgs84)
ext(scm_chm_45ft_wgs84) == ext(scm_aspect_45ft_resample_wgs84)
ext(scm_chm_45ft_wgs84) == ext(dist_to_rivers_wgs84)
ext(scm_chm_45ft_wgs84) == ext(dist_to_oceans_wgs84)
ext(scm_chm_45ft_wgs84) == ext(scm_postfire_chm_45ft_wgs84)

# resolution
res(scm_chm_45ft_wgs84) == res(scm_dem_45ft_resample_wgs84)
res(scm_chm_45ft_wgs84) == res(scm_aspect_45ft_resample_wgs84)
res(scm_chm_45ft_wgs84) == res(dist_to_rivers_wgs84)
res(scm_chm_45ft_wgs84) == res(dist_to_oceans_wgs84)
res(scm_chm_45ft_wgs84) == res(scm_postfire_chm_45ft_wgs84)
```

#### Save the rasters

These are the rasters that will be used for Maxent modeling in `Wallace`!

```{r}
writeRaster(scm_chm_45ft_wgs84, 
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/scm_chm_wgs84.tif", 
            overwrite = TRUE)
writeRaster(scm_dem_45ft_resample_wgs84,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/scm_dem_wgs84.tif",
            overwrite = TRUE)
writeRaster(scm_aspect_45ft_resample_wgs84,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/scm_aspect_wgs84.tif",
            overwrite = TRUE)
writeRaster(dist_to_rivers_wgs84,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/distance_to_rivers_wgs84.tif",
            overwrite = TRUE)
writeRaster(dist_to_oceans_wgs84,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/distance_to_oceans_wgs84.tif",
            overwrite = TRUE)
writeRaster(scm_postfire_chm_45ft_wgs84,
            "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/WGS_84/scm_postfire_chm_wgs84.tif",
            overwrite = TRUE)
```
