---
title: "Pacific Veg Map data processing"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Overview

-   This quarto doc serves to process all rasters downloaded from Pacific Veg Map to get them ready for Maxent modeling with the `Wallace` package

-   At the end of the processing, there will be a single raster file for the following forest structure & topographic rasters that covers the entire extent of the Santa Cruz Mountains:

-   Canopy height

- Post-fire canopy height 

-   Canopy cover

-   Digital Elevation Model (DEM)

-   Aspect

-   Slope

-   The single raster files were created by merging the rasters from Santa Cruz, Santa Clara, and San Mateo counties

-   All files from Pacific Veg Map, and all rasters for this analysis, will eventually have a resolution of 45 ft

## Load packages

```{r}
library(terra)
library(sf)
library(here)
library(tidyverse)
library(tmap)
# library(arcpullr)
```

## Load data

#### Forest structure

```{r}
#---------------- fine-scale vegetation ----------------#
cruz_clara_finescale_veg <- read_sf("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/cruz_clara/cruz_clara_finescale_veg/cruz_clara_finescale_veg.gdb")
crs(cruz_clara_finescale_veg)


#--------------- 3ft resolution canopy height models ---------------#

# santa cruz (4.6 GB)
santa_cruz_chm_3ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/santa_cruz/santa_cruz_CHM_res3ft_ft/SANTA_CRUZ_CHM_2020_3FT_COMBINED.tif")
crs(santa_cruz_chm_3ft)

# santa clara (11.9 GB)
santa_clara_chm_3ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/santa_clara/santa_clara_CHM_res3ft_ft/SANTA_CLARA_2020_CHM_3FT.tif")
crs(santa_clara_chm_3ft)

# san mateo (2.6 GB)
san_mateo_chm_3ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/san_mateo/san_mateo_CHM_res3ft_ft/SAN_MATEO_2017_CHM_3FT.tif")
crs(san_mateo_chm_3ft)

# CZU post-fire
santa_cruz_postfire_chm_3ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/cruz_clara/santa_cruz_postfire_CHM_res3ft_ft/CZU_POSTFIRE_CHM_3FEET.tif")
```

#### Topographic

```{r}
#---------------- digital elevation model (DEM) ----------------#

# santa cruz & santa clara (5m) 
santa_cruz_clara_dem_5m <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/raw_data/pacific_veg_map/cruz_clara/cruz_clara_dem_res5m_ft/ELEVATION_5M_UTM10_VALUES_FEET.tif")

# san mateo (20m)
san_mateo_dem_20m <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/san_mateo/san_mateo_dem_20m/san_mateo_dem_20m.tif")
res(san_mateo_dem_20m)

# make sure both have same CRS
crs(santa_cruz_clara_dem_5m) == crs(san_mateo_dem_20m)
```

## Change the resolution of vegetation data with `terra::aggregate()`

#### Check to make sure all rasters are in the same CRS

###### Canopy height model (CHM)

```{r}
# santa cruz & clara
if (crs(santa_cruz_chm_3ft) == crs(santa_clara_chm_3ft)){
  print("Cruz & Clara CHM rasters are in the same CRS")
} else {
  warning("Cruz & Clara CHM rasters are NOT in the same CRS")
}

# santa cruz & san mateo
if (crs(santa_cruz_chm_3ft) == crs(san_mateo_chm_3ft)){
  print("Cruz & Mateo CHM rasters are in the same CRS")
} else {
  warning("Cruz & Mateo CHM rasters are NOT in the same CRS")
}

# # project the San Mateo CHM to the same CRS as the Santa Cruz & Clara CHM
# san_mateo_chm_3ft <- project(san_mateo_chm_3ft, 
#                              crs(santa_cruz_chm_3ft))

# confirm they're in the same crs
if (crs(santa_cruz_chm_3ft) == crs(san_mateo_chm_3ft)){
  print("Cruz & Mateo CHM rasters are in the same CRS")
} else {
  warning("Cruz & Mateo CHM rasters are NOT in the same CRS")
}

# # save the projected San Mateo CHM
# san_mateo_chm_3ft_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/san_mateo/san_mateo_CHM_res3ft_ft_nad83_2011/san_mateo_CHM_res3ft_ft_nad83_2011.tif"
# writeRaster(san_mateo_chm_3ft, san_mateo_chm_3ft_file_path, 
#             overwrite = TRUE)
```

#### Change the resolution of the rasters from 3 ft to 45 ft

###### Canopy height model (CHM)

```{r}
#---------- change resolution of rasters ----------#

# santa cruz
santa_cruz_chm_45ft <- aggregate(santa_cruz_chm_3ft, 
                                 fact = 15, 
                                 fun = max) # CHM should be max height
#res(santa_cruz_chm_45ft) # QC

# santa clara
santa_clara_chm_45ft <- aggregate(santa_clara_chm_3ft, 
                                  fact = 15, 
                                  fun = max)
#res(santa_clara_chm_45ft) # QC

# san mateo
san_mateo_chm_45ft <- aggregate(san_mateo_chm_3ft, 
                                fact = 15, 
                                fun = max)
#res(san_mateo_chm_45ft) # QC


#---------- save the aggregated rasters ----------#

# santa cruz
cruz_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/santa_cruz/santa_cruz_CHM_res45ft_ft/santa_cruz_chm_45ft.tif"

writeRaster(santa_cruz_chm_45ft, cruz_file_path, 
            overwrite = TRUE)

# santa clara
clara_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/santa_clara/santa_clara_CHM_res45ft_ft/santa_clara_chm_45ft.tif"

writeRaster(santa_clara_chm_45ft, clara_file_path, 
            overwrite = TRUE)

# san mateo
mateo_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/san_mateo/san_mateo_CHM_res45ft_ft/san_mateo_chm_45ft.tif"

writeRaster(san_mateo_chm_45ft, mateo_file_path, 
            overwrite = TRUE)
```

## Change the resolution of topographic data with `terra::aggregate()`

#### Change the San Mateo DEM resolution to 5m

```{r}
# make sure the DEMs are in the same CRS
if (crs(santa_cruz_clara_dem_5m) == crs(san_mateo_dem_20m)){
  print("Cruz/Clara & Mateo DEM rasters are in the same CRS!")
} else {
  warning("Cruz/Clara & Mateo DEM rasters are NOT in the same CRS")
}


# use the terra::disaggregate() function to make the SM DEM 5m
san_mateo_dem_5m <- disagg(san_mateo_dem_20m, 
                           fact = 4, 
                           method = "bilinear")
# res(san_mateo_dem_5m) == res(santa_cruz_clara_dem_5m) # QC

# # save the 5m San Mateo raster 
# sm_5m_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/san_mateo/san_mateo_dem_5m/san_mateo_dem_5m.tif"
# writeRaster(san_mateo_dem_5m, sm_5m_file_path, 
#             overwrite = TRUE)

# merge the DEMs to have one DEM for entire SCM region
# order should be: (1) Santa Cruz; (2) Santa Clara; (3) San Mateo
# in areas where the rasters overlap, values of the first raster will be used
scm_dem_5m <- merge(santa_cruz_clara_dem_5m, san_mateo_dem_5m, 
                    first = TRUE)
```

#### Project the combined SCM DEM into the SCM CHM CRS

```{r}
# load the SCM 45 ft CHM 
scm_chm_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_CHM_res45ft_ft/scm_chm_45ft.tif")

# load the SCM 5m DEM 
scm_dem_5m <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_dem_5m/scm_dem_5m.tif")
crs(scm_dem_5m)
res(scm_dem_5m)

# load the SCM region boundary
scm_region_boundary <- read_sf("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_region_boundary/scm_region_boundary.shp")

# put the SCM region boundary in the same CRS as the DEM
scm_region_boundary <- st_transform(scm_region_boundary, 
                                    crs(scm_dem_5m))
crs(scm_region_boundary)

# clip and mask the DEM to the SCM region boundary
scm_dem_5m <- mask(scm_dem_5m, scm_region_boundary) %>% 
  crop(scm_region_boundary)
plot(scm_dem_5m)
```

#### Project the aspect raster to the same CRS as the DEM

```{r}
# load the aspect raster
scm_aspect_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_aspect_45ft/scm_aspect_45ft.tif")
crs(scm_aspect_45ft)

# load the DEM raster
scm_dem_45ft <- rast("C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_dem_45ft/scm_dem_45ft.tif")
crs(scm_dem_45ft)

crs(scm_aspect_45ft) == crs(scm_dem_45ft)

# project the aspect raster to the same CRS as the DEM
scm_aspect_45ft_proj <- project(scm_aspect_45ft, 
                                crs(scm_dem_45ft))

# confirm they're in the same CRS
crs(scm_aspect_45ft_proj) == crs(scm_dem_45ft)

# save the projected aspect raster
scm_aspect_45ft_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_aspect_45ft/scm_aspect_45ft.tif"

writeRaster(scm_aspect_45ft_proj, scm_aspect_45ft_file_path, 
            overwrite = TRUE)
```

```{r}
# put the DEM in the same CRS as the CHM
scm_dem_5m_proj <- project(scm_dem_5m, 
                           crs(scm_chm_45ft))

# confirm they're in the same CRS
crs(scm_chm_45ft) == crs(scm_dem_5m_proj)
```

#### Resample the SCM DEM to the SCM CHM resolution & position

```{r}
# resample the DEM to the CHM resolution
scm_dem_45ft <- resample(scm_dem_5m_proj, 
                         scm_chm_45ft, 
                         method = "bilinear")
res(scm_dem_45ft) == res(scm_chm_45ft) # QC

# save the resampled DEM
scm_dem_45ft_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_dem_45ft/scm_dem_45ft.tif"

writeRaster(scm_dem_45ft, scm_dem_45ft_file_path, 
            overwrite = TRUE)
```

## Merge the aggregated rasters

#### Canopy height model (CHM)

```{r}
# load the aggregated (45 ft resolution) CHM rasters
cruz_chm_45ft <- rast(cruz_file_path)
clara_chm_45ft <- rast(clara_file_path)
mateo_chm_45ft <- rast(mateo_file_path)
# make sure they're in the same CRS
crs(cruz_chm_45ft) == crs(clara_chm_45ft)
crs(clara_chm_45ft) == crs(mateo_chm_45ft)


# merge the rasters to have one CHM for entire SCM region
# order should be: (1) Santa Cruz; (2) Santa Clara; (3) San Mateo
# in areas where the rasters overlap, values of the first raster will be used
scm_chm_45ft <- merge(cruz_chm_45ft, clara_chm_45ft, mateo_chm_45ft, 
                      first = TRUE)

# plot with tmap to QC
tm_shape(scm_chm_45ft) +
  tm_raster() + 
  tm_layout(legend.outside = TRUE)

# save the merged CHM!
scm_chm_45ft_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/pacific_veg_map/scm_region_combined/scm_CHM_res45ft_ft/scm_chm_45ft.tif"

writeRaster(scm_chm_45ft, scm_chm_45ft_file_path, 
            overwrite = TRUE)
```

#### Mask the CHM to SCM region

```{r}
# put the SCM region boundary in the same CRS as the CHM
scm_region_boundary <- st_transform(scm_region_boundary, 
                                    crs(scm_chm_45ft))
crs(scm_region_boundary) == crs(scm_chm_45ft)

# mask/crop the CHM 
scm_chm_45ft <- mask(scm_chm_45ft, scm_region_boundary) %>% 
  crop(scm_region_boundary)
plot(scm_chm_45ft)


# save the new CHM
scm_chm_45ft_file_path <- "C:/Users/maxpe/Documents/Bren Courses/ESM 270p/data/new_data/wallace_env_data/scm_chm_45ft.tif"

writeRaster(scm_chm_45ft, scm_chm_45ft_file_path, 
            overwrite = TRUE)
```

#### Digital Elevation Models (DEMs)

```{r}
# did this with the Mosaic to Raster tool in ArcGIS Pro because R kept crashing
```
